---
output: html_document
title: 'Gamma Gamma Hey (an options analysis) Part Two:'
layout: post
---

# Building Basic Options SME


[Github link](https://github.com/sebidelamata/optionsBacktesting){target="_blank"}

## Abstract

In this post we will use data collected from [post one](https://sebidelamata.github.io/datascienceblog/2022/06/12/Gamma-Gamma-Hey-Part-One.html){target="_blank"} to educate ourselves on the basics behind option valuation as well as some common strategies used in option trading. To accomplish this we will be creating a series of functions as well as visualize some real world trades with Plotly. 

## Single Option Valuation

Equity options generally serve as a contract giving the owner the right to sell (a put) or buy (a call) 100 shares of the underlying at a given price. We will start of modeling these as European style options to start for the sake of simplicity. European options can be exercised at the date of expiration, where American style options allow for early exercise up to and including the date of expiration.

### Call Options

As stated before, equity call options give the owner the right buy 100 shares of the underlying at a given price. The value of this contract has a convex payoff structure compared to the underlying. The contract will be worth nothing if it expires before the underlying price has surpassed the strike price of the contract. It will otherwise be worth the difference between the market price of the underlying and the strike price of the contract. This basic payoff function can be modeled like so:

```{r}

library(dplyr)

# The payoff for a call buyer at expiration date T is given by 
# max(0, ST – X)
call_buyer_payoff <- function(underlying_price_expiration, strike_price){
  
  # either zero or the difference in the strike and expiration
  max(0, underlying_price_expiration - strike_price)
  
}

```

We can model the call buyer payoff for an at-the-money QQQ Jan 20th, 2023 call if we had bought the option on Friday, July 1st 2022.

```{r}

library(RPostgreSQL)
library(getPass)

# set driver name
driver_name <- dbDriver(drvName = "PostgreSQL")

# establish database connection
db <- DBI::dbConnect(driver_name,
                     dbname = "sebi",
                     host = "192.168.0.12",
                     port = 5432,
                     user = "sebi",
                     password = getPass("Enter Password:")
)

# our query
sql_query_string <- "SELECT * FROM watchlist_data WHERE scrape_date = '2022-07-01' AND underlying_ticker = 'QQQ' AND contract_type = 'C' AND expiration_date = '2023-01-20';"

# combination of our connnection stuff and our query stuff
res <- dbSendQuery(db, sql_query_string)

# fetch our data query from the database and return a dataframe
data_pull <- dbFetch(res)

# clear our query results
dbClearResult(res)

# politely disconnect from the database
dbDisconnect(db)

# select the at the money
qqq_atm_call_jan_twenty <- data_pull[data_pull$strike_price == 280,]

two_fifty_call_payoff <- sapply(
  seq(250,400),
  FUN = function(x){
    call_buyer_payoff(
      underlying_price_expiration = x, 
      strike_price = qqq_atm_call_jan_twenty$strike_price
      )
    }
)

# create a df to hold values with possible strikes at expiration
two_fifty_call_payoff <- data.frame(
  possible_profits = two_fifty_call_payoff,
  possible_expiration_prices = seq(250, 400)
)

# create a quick function to plot our payouts (we'll be doing a few)
# plotting our payoff functions (plotly)
# takes a profit df (2 columns profits and strikes)
# and a Title for the plot
lineplot_profit <- function(profit_df, title){
  
  library(ggplot2)
  library(plotly)
  library(scales)
  library(wesanderson)
  
  profit_lineplot <- ggplot(
    data = profit_df,
    aes(
      x = possible_expiration_prices,
      y = possible_profits
    )
  ) +
    geom_line(color = wesanderson::wes_palette("Royal1")[3]) +
    ggtitle(title) +
    xlab("Possible Expiration Prices") +
    ylab("Profit (USD $)") +
    theme_minimal() +
    theme(
      panel.background = element_rect(fill = wes_palette("Darjeeling2")[5], color = wes_palette("Darjeeling2")[5]),
      plot.background = element_rect(fill = wes_palette("Darjeeling2")[5], color = wes_palette("Darjeeling2")[5]),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(color = wesanderson::wes_palette("Royal1")[3]),
      axis.text.y = element_text(color = wesanderson::wes_palette("Royal1")[3]),
      legend.position = "none",
      text = element_text(color = wesanderson::wes_palette("Royal1")[3])
      ) +
    scale_y_continuous(
      labels = scales::dollar_format(),
      limits = c(
        min(profit_df$possible_profits) - abs(0.2 * min(profit_df$possible_profits)),
        max(profit_df$possible_profits) + abs(0.2 * max(profit_df$possible_profits))
        )
      ) +
    scale_x_continuous(labels = scales::dollar_format()) +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = wesanderson::wes_palette("Royal1")[3]
      ) +
    scale_color_manual(values = wesanderson::wes_palette("Royal1")[3])
  
  profit_lineplot <- plotly::ggplotly(profit_lineplot)
                      
  return(profit_lineplot)
  
}

# now let's plot our payout
two_fifty_call_payoff <- lineplot_profit(
  two_fifty_call_payoff, 
  "Basic payoff, Jan 20th, 2023 280 Call"
  )


two_fifty_call_payoff

```


